---
title: 'Survival analysis'
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
if (!require(survival)) {install.packages("survival")}
if (!require(ggplot2)) {install.packages("ggplot2")}
if (!require(dplyr)) {install.packages("dplyr")}
if (!require(gridExtra)) {install.packages("gridExtra")}
if (!require(survminer)) {install.packages("survminer")}
if (!require(coin)) {install.packages("coin")}
if (!require(ggfortify)) {install.packages("ggfortify")}
```

# Data set description

Data set depicts survival in a randomised trial comparing two treatments for ovarian cancer.

Following variables are present:

\      **futime**:	survival or censoring time

\      **fustat**:	censoring status

\      **age**:  in years

\      **resid.ds**:   residual disease present (1=no,2=yes)

\      **rx**:	treatment group

\      **ecog.ps**:  [ECOG performance status](https://ecog-acrin.org/resources/ecog-performance-status)

# Exploratory data analysis

Data has following structure:

```{r comment="", echo=FALSE}
str(ovarian)
```

Variables **fustat**, **resid.ds**, **rx** and **ecog.ps** can be converted to factors.

```{r echo=FALSE, include=FALSE}
ova <- ovarian
ova$rx <- factor(ova$rx)
ova$resid.ds <- factor(ova$resid.ds)
ova$ecog.ps <- factor(ova$ecog.ps)
ova$age_groups <- factor(ifelse((ova$age < 57), "Younger than 57", "Older than 57"))
```

Data has only **26** observations.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ova, aes(x = age, y = futime, col=factor(rx))) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(x = "Age", y = "Survival time") +
    scale_color_discrete(name = "Treatment group") +
    theme(plot.title = element_text(face="bold", size=16, hjust = 0.5),
        axis.title.y = element_text(size=14, vjust = 0.5),
        axis.title.x = element_text(size=14, vjust = 0.5))
cortest <- cor.test(ova$futime, ova$age)
```

Variables **age** and **futime** are negatively correlated (**r = `r cortest$estimate`, p-value = `r sprintf("%.3g", cortest$p.value)`**)

**`r sum(ova$fustat == 1)`** of **`r length(ova$fustat)`** observations are censored.

Groups formed by categorical variables are badly balanced, probably due to lack of observations.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ova %>% group_by(fustat, resid.ds, rx, ecog.ps) %>% summarise(n=n())
```

# Survival analysis

## Kaplan–Meier curves

First, we need to build model using **fustat** and **futime** variables. This will describe changes in survival chances depending on time.

```{r echo=FALSE, comment=""}
cb_palete <- rev(c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"))
km_fit <- survfit(Surv(futime, fustat) ~ 1, data=ova)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit, surv.colour = cb_palete[3], xlab = "Time", ylab = "Survival rate")
```

We can see very large CIs. Again, this may be caused by insufficiency of observations. However this model yields a little information, we can analyze survival rate depending on other factors.

<hr>

### Survival rate depending on ECOG status

```{r echo=FALSE, comment=""}
km_fit <- survfit(Surv(futime, fustat) ~ ecog.ps, data=ova)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit, xlab = "Time", ylab = "Survival rate")
logrank_test(Surv(futime, fustat) ~ ecog.ps, data=ova)
pairwise_survdiff(Surv(futime, fustat) ~ ecog.ps, data=ova, p.adjust.method = "BH")
```
Survival rate does not significantly differ between ECOG status 1 and 2.

<hr>

### Survival rate depending on treatment

```{r echo=FALSE, comment=""}
km_fit <- survfit(Surv(futime, fustat) ~ rx, data=ova)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit, xlab = "Time", ylab = "Survival rate")
logrank_test(Surv(futime, fustat) ~ rx, data=ova)
```

Survival rate does not significantly differ between different treatment types.

<hr>

### Survival rate depending on residual disease presence

```{r echo=FALSE, comment=""}
km_fit <- survfit(Surv(futime, fustat) ~ resid.ds, data=ova)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit, xlab = "Time", ylab = "Survival rate")
logrank_test(Surv(futime, fustat) ~ resid.ds, data=ova)
```
Survival rate does not significantly differ depending on residual disease presence.

<hr>

### Survival rate depending on age group

```{r echo=FALSE, comment=""}
km_fit <- survfit(Surv(futime, fustat) ~ age_groups, data=ova)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit, xlab = "Time", ylab = "Survival rate")
logrank_test(Surv(futime, fustat) ~ age_groups, data=ova)
```

Survival rate does not significantly differ depending on age groups.

<hr>

### Survival rate depending on age group and treatment

```{r echo=FALSE, comment=""}
ova$agerx <- factor(paste(ova$age_groups, ova$rx, sep="_"))
km_fit <- survfit(Surv(futime, fustat) ~ agerx, data=ova)
# summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit, xlab = "Time", ylab = "Survival rate")
pairwise_survdiff(Surv(futime, fustat) ~ agerx, data=ova, p.adjust.method = "bonferroni", rho=1)
```

Here **peto & peto logrank test** was made and bonferroni adjustment was applied, because now we have 4 groups to compare. We can notice that type 1 treated people in different age groups significantly differ in survival rates.

<hr>

### Survival rate depending on residual disease presence and treatment

```{r echo=FALSE, comment=""}
ova$residrx <- factor(paste(ova$resid.ds, ova$rx, sep="_"))
km_fit <- survfit(Surv(futime, fustat) ~ residrx, data=ova)
# summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit, xlab = "Time", ylab = "Survival rate")
pairwise_survdiff(Surv(futime, fustat) ~ residrx, data=ova, p.adjust.method = "bonferroni", rho=1)
```

According to logrank test with bonferroni adjustment these groups do not have any significant differences in survival rate.

## Analysis of hazard factors

### Proportional hazards regression model

To estimate hazard factors Cox model can be built.

```{r echo=FALSE, comment=""}
cox <- coxph(Surv(futime, fustat) ~ resid.ds + ecog.ps + rx + age_groups, data = ova)
summary(cox)
```

<a id="COX2"></a>
This model does not yield any covariates. This could happen because **age** variable directly depends on time and Cox model assumes that it is not. So we can remove this variable from our model.

```{r echo=FALSE, comment=""}
cox <- coxph(Surv(futime, fustat) ~ resid.ds + ecog.ps + rx, data = ova)
summary(cox)
```

Now we can see that residual disease presence became significant hazard factor. However we need to check if it is not changing in time.

To do that we can build Aalen model and visualize it:

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=""}
aa_fit <- aareg(Surv(futime, fustat) ~ resid.ds + ecog.ps + rx, data = ova)
autoplot(aa_fit)
```

On this chart we can see that **resid.ds2** changes over time and thus should be interpreted with caution.

### Hazard ratio

In order to characterize differences in survival rates numerically we can use hazard ratios (HRs).

```{r message=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(futime, fustat) ~ resid.ds + ecog.ps + rx, data = ova)
ggforest(fit.coxph, data = ova)
```

Here we can see that **resid.ds2** HR significantly differ from 1. In fact, it is the same thing that we obtained [here](#COX2). HR for **resid.ds2** relatively to reference is 3.84. It means that **people with residual disease presence have 76.8% ($\frac{3.84}{1+3.85}$) higher chance of earlier status change (we do not really know what does *pustat* levels mean)**.

# Conclusion

Here we made a survival analysis of ovarian data set.

We showed that:

+ Type 1 treated people below age of 57 have lower survival rates than people older than 57.

+ Residual disease presence might be significant hazard factor.

+ Survival rates among other groups and inside them do not significantly differ.












